---
output:
  md_document:
    variant: gfm
---

# deepnnet

`deepnnet` fits multilayer perceptrons for tabular classification and regression in
R using `torch`. It is designed for formula-based workflows: pass a formula and data frame,
let the package preprocess numeric and categorical predictors, then use
standard prediction, metrics, tuning, and interpretation helpers.

Source repository: <https://github.com/ielbadisy/mlp>

`deepnnet` is not yet on CRAN. Install it from GitHub with:

```r
remotes::install_github("ielbadisy/mlp")
```

<!-- CRAN first submission: keep the GitHub install instructions until the package is on CRAN. -->

For local development without reinstalling after each edit:

```r
pkgload::load_all(".")
```

Do not source individual files such as `R/mlp.R`; the exported functions depend
on helpers that are loaded through the package namespace.

## Features

- Binary classification, multiclass classification, and regression.
- Formula-based interface.
- Automatic preprocessing for missing numeric values, missing categorical
  values, and unseen prediction-time categories.
- Configurable hidden layers, activation, dropout, batch normalization,
  residual blocks, gated blocks, and input projection.
- Early stopping, learning-rate schedules, and common `torch` optimizers.
- Task-aware predictions, metrics, training-history plots, tuning, and
  permutation importance.

## Classification Example

```{r classification, echo=TRUE, message=FALSE, warning=FALSE}
library(deepnnet)

fit <- mlp(
  Species ~ .,
  data = iris,
  epochs = 10,
  validation = 0.2,
  verbose = TRUE,
  seed = 1
)

predict(fit, iris[1:5, ], type = "class")
predict(fit, iris[1:5, ], type = "prob")
pred <- predict(fit, iris, type = "class")
mlp_metrics(iris$Species, pred, task = "classification")
```

## Regression Example

```{r regression, echo=TRUE, message=FALSE, warning=FALSE}
fit_reg <- mlp(
  mpg ~ disp + hp + wt,
  data = mtcars,
  task = "regression",
  epochs = 10,
  validation = 0.2,
  verbose = FALSE,
  seed = 2
)

pred_reg <- predict(fit_reg, mtcars, type = "response")
mlp_metrics(mtcars$mpg, pred_reg, task = "regression")
```

## Tuning

```{r tuning, echo=TRUE, message=FALSE, warning=FALSE}
tuned <- tune_mlp(
  Species ~ .,
  data = iris,
  grid = list(
    hidden_units = list(c(8), c(16, 8)),
    activation = c("relu"),
    dropout = c(0, 0.1),
    lr = c(1e-3),
    epochs = c(10)
  ),
  patience = 3,
  seed = 1,
  verbose = FALSE
)

tuned$results
tuned$best_fit
```

## Interpretation

```{r interpretation, echo=TRUE, message=FALSE, warning=FALSE}
plot_history(fit)
ggplot2::autoplot(fit)

importance <- perm_importance(fit, iris[, -5], iris$Species)
plot(importance)
```

## Status

Version `0.5.0` supports binary classification, multiclass classification, and
regression for tabular data frames with automatic preprocessing.
