---
output:
  md_document:
    variant: gfm-smart
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# densemlp

`densemlp` provides dense feed-forward neural network models (multilayer
perceptrons) for tabular **regression**, **classification** and **survival
analysis** in R. Forward propagation, backpropagation and Adam optimization
are implemented natively in C++ via `RcppArmadillo`, with hand-derived
closed-form gradients: there is **no `torch` / `libtorch` dependency**, so
the package installs and trains without a separate deep-learning runtime.

The optional extensions - dropout, batch normalization, residual
connections, gated blocks, a learned cross-feature interaction layer, a
linear input projection, exponential moving-average weights, learning-rate
schedules and internal bootstrap ensembles - enrich the architecture
without changing its model class: the models stay dense feed-forward neural
networks.

`task` is optional. When set to `"auto"` (the default) the task is inferred
from the outcome: numeric to `"regression"`, a factor/character to
`"binary"` / `"multiclass"`, a `survival::Surv()` object to `"survival"`.

## Installation

```r
# CRAN
install.packages("densemlp")

# development version
remotes::install_github("ielbadisy/densemlp")
```

## Regression

```{r regression, message=FALSE, warning=FALSE}
library(densemlp)

set.seed(1)
n <- 300
x <- data.frame(a = rnorm(n), b = rnorm(n), c = rnorm(n))
y <- 2 * x$a - x$b + rnorm(n, sd = 0.2)

fit <- densemlp(x, y, hidden_units = c(32, 16), epochs = 60, seed = 1)
pred <- predict(fit, x)
densemlp_metrics(y, pred, task = "regression")
```

## Classification

```{r classification, message=FALSE, warning=FALSE}
fit <- densemlp(Species ~ ., data = iris, epochs = 40, seed = 1)
predict(fit, iris[c(1, 60, 120), ], type = "prob")
predict(fit, iris[c(1, 60, 120), ], type = "class")
```

## Survival

```{r survival, message=FALSE, warning=FALSE}
library(survival)
data(lung, package = "survival")
lung <- na.omit(lung[, c("time", "status", "age", "sex", "ph.ecog")])
sy <- Surv(lung$time, lung$status == 2)
sx <- lung[, c("age", "sex", "ph.ecog")]

sfit <- densemlp(sx, sy, task = "survival", hidden_units = c(16, 8), epochs = 60, seed = 1)
densemlp_metrics(sy, predict(sfit, sx, type = "response"), task = "survival")
```

## More

* `cv_densemlp()` - k-fold cross-validation.
* `tune_densemlp()` - grid search over the architecture and optimizer knobs.
* `perm_importance()` - model-agnostic permutation importance.
* `plot_history()` / `plot()` - training and validation loss curves.

See `?densemlp` and `vignette("densemlp-intro")` for the full API.

## Citation

```{r citation}
citation("densemlp")
```
