---
title: "Getting Started with deepnnet"
output:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with deepnnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r}
library(deepnnet)
```

`mlp` trains multilayer perceptrons for tabular data. It accepts a formula and data frame, preprocesses numeric and categorical
predictors, and returns an `mlp_fit` object with methods for prediction,
plotting, metrics, tuning, and variable importance.

This vignette documents version `0.5.0`.

This vignette assumes the package is installed or loaded with
`pkgload::load_all(".")`. Do not source individual files such as `R/mlp.R`
because exported functions rely on helpers loaded through the package
namespace.

## Classification

For classification, use a factor outcome. `task = "auto"` infers the task from
the outcome type, so it can usually be omitted.

```{r classification-fit}
fit <- mlp(
  Species ~ .,
  data = iris,
  epochs = 10,
  patience = 3,
  verbose = FALSE,
  seed = 1
)

fit
```

Class predictions are returned as factors with the same levels as the training
outcome.

```{r classification-predictions}
predict(fit, iris[1:5, ], type = "class")
round(predict(fit, iris[1:5, ], type = "prob"), 3)
```

Use task-aware metrics to summarize predictions.

```{r classification-metrics}
pred <- predict(fit, iris, type = "class")
mlp_metrics(iris$Species, pred, task = "classification")
```

## Regression

For regression, use a numeric outcome or set `task = "regression"` explicitly.

```{r regression-fit}
fit_reg <- mlp(
  mpg ~ disp + hp + wt,
  data = mtcars,
  task = "regression",
  epochs = 10,
  patience = 3,
  verbose = FALSE,
  seed = 2
)

fit_reg
```

```{r regression-predictions}
pred_reg <- predict(fit_reg, mtcars, type = "response")
head(round(pred_reg, 2))
mlp_metrics(mtcars$mpg, pred_reg, task = "regression")
```

## Training Diagnostics

The fitted object stores training history, which can be displayed directly or
through the `ggplot2::autoplot()` method.

```{r plot-history}
plot_history(fit)
```

```{r autoplot-history}
ggplot2::autoplot(fit)
```

## Tuning

`tune_mlp()` evaluates a grid of hyperparameters and, by default, refits the
best configuration on the supplied data.

```{r tuning}
tuned <- tune_mlp(
  Species ~ .,
  data = iris,
  grid = list(
    hidden_units = list(c(8), c(16, 8)),
    activation = c("relu"),
    dropout = c(0),
    batch_size = c(8),
    lr = c(1e-3),
    epochs = c(10)
  ),
  patience = 3,
  seed = 3,
  verbose = FALSE
)

tuned$results
```

## Permutation Importance

Permutation importance estimates how much a metric changes after shuffling each
predictor.

```{r permutation-importance}
importance <- perm_importance(fit, iris[, -5], iris$Species)
importance$data
plot(importance)
```
