---
title: "Getting Started with densemlp"
output:
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with densemlp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r}
library(densemlp)
```

`densemlp` trains dense feed-forward multilayer perceptrons for tabular data.
It accepts a formula and data frame, preprocesses numeric and categorical
predictors, and returns a `densemlp_fit` object with methods for prediction,
plotting, metrics, tuning, and variable importance.

The models implemented in `densemlp` are dense feed-forward multilayer
perceptrons. Optional components such as dropout, batch normalization,
residual connections, gated blocks, and input projection extend the basic
dense multilayer perceptron architecture but do not change the model class
into a convolutional, recurrent, transformer, or tree-based model.

`task` is optional in the main API. When it is set to `"auto"` or omitted, the
task is inferred from the outcome. Use it only when you need to override that
inference.

This vignette documents version `0.5.0`.

This vignette assumes the package is installed or loaded with
`pkgload::load_all(".")`. Do not source individual files such as `R/densemlp.R`
because exported functions rely on helpers loaded through the package
namespace.

`densemlp` implements dense feed-forward multilayer perceptrons for tabular
data. Each hidden block is centered on a fully connected transformation,
optionally combined with activation functions, dropout, batch normalization,
residual connections, gated mechanisms, and input projection.

## Classification

For classification, use a factor outcome. The task is inferred from the
outcome type, so `task` can usually be omitted.

```{r classification-fit}
fit <- densemlp(
  Species ~ .,
  data = iris,
  epochs = 10,
  patience = 3,
  verbose = FALSE,
  seed = 1
)

fit
```

Class predictions are returned as factors with the same levels as the training
outcome.

```{r classification-predictions}
predict(fit, iris[1:5, ], type = "class")
round(predict(fit, iris[1:5, ], type = "prob"), 3)
```

Use task-aware metrics to summarize predictions.

```{r classification-metrics}
pred <- predict(fit, iris, type = "class")
densemlp_metrics(iris$Species, pred)
```

## Regression

For regression, use a numeric outcome. The task is inferred automatically for
numeric outcomes.

```{r regression-fit}
fit_reg <- densemlp(
  mpg ~ disp + hp + wt,
  data = mtcars,
  epochs = 10,
  patience = 3,
  verbose = FALSE,
  seed = 2
)

fit_reg
```

```{r regression-predictions}
pred_reg <- predict(fit_reg, mtcars, type = "response")
head(round(pred_reg, 2))
densemlp_metrics(mtcars$mpg, pred_reg)
```

## Training Diagnostics

The fitted object stores training history, which can be displayed directly or
through the `ggplot2::autoplot()` method.

```{r plot-history}
plot_history(fit)
```

```{r autoplot-history}
ggplot2::autoplot(fit)
```

## Tuning

`tune_densemlp()` evaluates a grid of hyperparameters and, by default, refits the
best configuration on the supplied data.

```{r tuning}
tuned <- tune_densemlp(
  Species ~ .,
  data = iris,
  grid = list(
    hidden_units = list(c(8), c(16, 8)),
    activation = c("relu"),
    dropout = c(0),
    batch_size = c(8),
    lr = c(1e-3),
    epochs = c(10)
  ),
  patience = 3,
  seed = 3,
  verbose = FALSE
)

tuned$results
```

## Permutation Importance

Permutation importance estimates how much a metric changes after shuffling each
predictor.

```{r permutation-importance}
importance <- perm_importance(fit, iris[, -5], iris$Species)
importance$data
plot(importance)
```
